Unifying the communicable disease spreading paradigm with Gompertzian growth

### 2022-11-28 ResponseToReviewers.txt

Response to reviewer #3:

Reviewer #3: After reading the article reviewed, I came to the following conclusions: from a mathematical point of view, the model seems to be correct (unless I have overlooked a mistake), while the interpretation of the results obtained is, to put it mildly, controversial, and in some respects incorrect or requiring additional explanations. In particular:

p.4. (after Eq.7) …At this limit the relative growth rate approaches infinity as Y→0, which seems odd under the hypothesis that the pathogen has just started spreading.
Comment: The form of the growth functions used in the biological domain excludes the possibility of reaching the limiting value Y=0, which is Y(t→0)=Y0. Therefore, the growth rate will never reach an infinite value.

Answer: This is a good point and the sentence has been removed.

p.6. …all entities are connected and communicating instantaneously as shown by Mombach et al.
Comment: Mombach et al. derived the Gompertz function in the framework of the cellular mean field theory, in which the presence of long-range (slowly decaying) interactions is taken into account. In biological systems, interactions at the distance (nonlocal correlations) even when they seem instantaneous, they do not allow the superluminal communication (information transfer), which, according to the theory of relativity is impossible.

Answer: I see that the the word "instantaneous" is inaccurate in this context and have changed this paragraph to 
"The SIRD model above also considers average macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \cite{smilkov2014beyond} which does not consider network effects explicitly. Rather, all entities are connected and communicating *through long-range interactions* as shown by Mombach et al. (2002) \cite{mombach2002mean}.
"

The idea was to quote the extra assumptions that are present in mean field theories which do not have a physical basis in the communicable disease paradigm, namely that pathogens have long-range interactions, which by construction would imply near-luminal effect-velocities, which is a clear violation of this paradigm.

p.6 …the biosphere was perturbed by an instantaneous external stressor, initiating a stress response to eventually bring mortality rates back to stability.
Comment: The authors should explain: what is the nature and origin of this globally affecting biosphere stressor, which plays a key role in the interpretation of the results obtained. The context of the work suggests that it plays a similar role to the cosmological constant in the general relativity, guaranteeing the stability (stationarity) of the solutions.

Answer: Here we have intentionally left out a discussion of such stressor mechanisms and causes due to the uncertainty of such speculations. We do not wish to derail an otherwise scientific investigation and in some ways falsification of the communicable disease paradigm into somewhat speculative alternative theories. However, for now we have tentatively added the following
[PLACEHOLDER: This stressor could be in the form of anomalous geomagnetic storms, or man-made disturbances in the atmosphere. For example, the ionosphere, which is the reservoir of the earth's electromagnetic potential, could have been disturbed either by man-made or natural events, causing organisms to experience an immune response. The exact mechanism of such hypotheses would need further study.]

More litterature review would need to be done to back up these claims.

p.7 …Comparing the stochastic stressor σ in Eq. 14 with the deterministic growth equation in Eq. 1 gives γ= σ/2…
Comment: Why in comparison σ1/2dW(t)/dt (delta-correlated Wiener component) is not included? Only the mean of this term is equal zero, whereas the compared Eqs.1 and 14 do not contain means.

By virtue of Eq. 1 being deterministic, it is implicitly interpreted as a mean (non-random) quantity. In fact, we make the argument that it is more like the median (not mean) of the miscroscopic equation in 14. I see that this was not clearly written and have appended an extra sentence to make this clear:
"This comparison is clear after taking the median of Eq. \ref{eq:microscopicOriginal} wherein the cross-term of $F(t)dW(t)$ goes to zero due to independence."

p.8 …We can now further appreciate the Richards parametrization as a transition from non-collaborative to collaborative growth as Θ→0.
Comment: In the limit Θ→0 the Richards (generalized logistic) model reduces to the Gompertz one. The latter function is solution of the non-local equation for the time-dependent More oscillator in the dissociation state (Phys. Rev. E 68(2), 021916, 2003), hence, it represents the coherent in space (collaborative) growth. In a similar manner one may prove that the Richards function is solution of the identical equation but for hyperbolic Rosen-Morse potential, therefore it also describes the coherent growth. Hence, both models are collaborative in the nature.

Answer: Thanks for pointing this out. I see there are nuances here that can easily be misinterpreted. My assumption is that when theta is 1,  we get the logistic equation and the growth is non-collaborative, while when theta is 0 we get a "maximally" coherent growth in line with the Morse oscillator. Based on the work of Petroni et al (https://arxiv.org/pdf/2004.10478.pdf) it seems logistic is not collaborative in the sense that "resources are not shared" to use their phrase. Based on this insight, I have modified the inaccurate statement above to:

"We can now further appreciate the Richards parametrization as a transition from non-collaborative to maximally collaborative growth as $\theta \rightarrow 0$. " By inserting the word "maximally", it is hopefully clear that we are on a continuoum of collaborative growth if we could allow such a concept.

p.11…In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states, and that the perceived pathogen travels at infinite speed throughout the population, rejecting the possibility of a disease propagation stage through a perceived transmission network.
Comment: This conclusion is inconsistent with the Feinberg's extended relativity (Phys. Rev. 159, 1089, 1967). According to Feinberg's theory, pathogens (observable, i.e. subluminal entities) cannot move faster than the speed of light c in a vacuum, because subluminal objects reaching infinite speed would cross the light barrier, which is impossible from the relativistic point of view. Non-local objects (fields) can move (propagate) with the allowed group velocity v=(c, ∞), but this kinetical characteristic does not apply to pathogens moving at v=(0,c). This part of the work should be removed or its interpretation changed.

Your point is well taken, and we of course agree that pathogens cannot travel at superluminal speeds. This is again a flaw in our communication. The intention is to present the implausibility of the tasks of the "pathogen", thus showing that no such pathogen could exist (near-lunumal speed and driven by susceptible instead of infected). We failed to statethe latter part clearly, namely that due to the implications of the gompertz model, we have to reject the concept of a pathogen in this context. A new theory driven by a non-local field must is our suggested replacement, but we are open to ideas. The whole conclusion is now re-written with the relevant section stating:

"
In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states. In this vein, the Richards parameter, $\theta$, is related although not identical to the number of higher-order interactions with the susceptible and the infected population in an augmented SIR model \cite{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$ and where the source of growth is driven by the number of susceptible and not the number of infected.
[..]
Observing Gompertzian growth through the communicable diseases paradigm, the perceived pathogen would have to travel at near-luminal speeds throughout the population, exemplified by the immediate decrease in the relative growth rate of the infected, and be driven by the susceptible as opposed to the infected. We are thus based on physical grounds forced to reject the concept of a pathogen in the context of Gompertzian growth. 
"

Additional remarks: A comparison of the SIRD with Gompertz model, graphically presented in Fig.1 does not reflect the differences of the models considered in an acceptable way. In order to compare the models, I suggest to calculate the basic reproduction number R0, which indicates how many secondary infections are caused by an infected individual at the beginning of an epidemic when infection spreading is of exponential type. Indispensable in the calculations formulae and data one may find in the Kong et al. work (PLoS One 16, e0252373, 2021). References [12], [48], [51] are incomplete and should be supplemented with the missing information.

In conclusion: the work NODY-D-22-02727 submitted for publication in Nonlinear Dynamics requires the inclusion of the corrections suggested, explanations mentioned and supplementary calculations that will reveal the advantages of the proposed model over the models functioning in the area of current research. Only then it will be possible to decide whether the work is suitable for publication or not.

### 2022-11-29 arxiv/main.tex
resond to reviewer 2

\insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of deceased from the Coronavirus, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing an SIRD model (green) with a Gompertz model (yellow) for a variety of countries in the period Jan-May 2020. Both models are fit using non-linear least squares according to equations \ref{eq:GompertzODE} and \ref{eq:SIRD_D} in the text. Although the Gompertz curve can also be obtained with a simple linear fit using Eq. \ref{eq:GompertzODE2param}, it is here obtained using non-linear least squares to put both models on equal footing. The temporal evolution of the SIRD model is obtained from the Runge-Kutta algorithm, while the Gompertz has a closed form for its temporal evolution. Each plot is annotated with an inferred transmission rate in the SIRD model with 1 standard deviation in brackets, assuming bi-normally distributed parameter estimates. Fitting is done with Python-Scipy and a 6-day moving average of deaths. Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \cite{dong2020interactive}.}

@@

\section*{Unifying the SIR model and the Gompertz model}
\subsection*{Comparing microscopic logistic and Gompertz growth}
The remarkable observation that the log-transformed domain is the natural one merits closer study. First, juxtapose the logistic model with the Gompertz model,

@@

\subsection*{Unification through a modified SIR model}
We can now reconcile the SIR model with the Gompertz curve. 
Inspired by the observation that in the linearized approximation there are only two coupled states, infected and susceptible, we augment the interaction term to higher orders. 
Then, we reverse the causality where the population of infected is now dependent on the population of the susceptible instead of the other way around as in Eq. \ref{eq:networkSIR},
\begin{equation}
\label{eq:reverseNetworkSIR}
\frac{d x_i}{dt} = \beta x_i\sum_j{a_{ij}}(1-x_j) \quad \forall i.
\end{equation} 
This crucial change is based on the environmental stressor hypothesis rather than a communicable disease assumption.
Furthermore, we will for the sake of simplicity assume all nodes in the network have exactly one neighbor and that there exists a unique path between all nodes,
\begin{equation}
\sum_i a_{ij} = 1 \quad \forall j.
\end{equation} 
If we let $s_i = 1 - x_i$ be the probability of being susceptible, the augmented and causality-reversed SIR model's infection term becomes 
\begin{equation}
\frac{d x_i}{dt} = \beta x_i\sum_j{a_{ij}}(s_{j} + s^2_j/2 + ...) - \alpha x_i\quad \forall i,
\end{equation}

@@

\begin{equation}
\frac{d x_i}{dt} = -\beta x_i\sum_j{a_{ij}}\ln{(1 - s_j)} - \alpha = -\beta x_i\sum_j{a_{ij}}\ln{x_j} - \alpha x_i \quad \forall i,
\end{equation}
As we are interested in the aggregate macroscopic behavior, we take averages in the log-domain, exploit our setup where $\mathbf{A}$ is a single mapping from one node to another, and simplify to
\begin{equation}
\frac{d}{{dt}} \ln\left[\prod_i^N{x_i}\right]^{\frac{1}{N}} = -\beta\ln\left[\prod_i^N{x_i}\right]^{\frac{1}{N}} - \frac{\alpha}{N}\sum_i^N x_i.
\end{equation}
If $\alpha\rightarrow 0$, then this equation will exhibit Gompertzian growth in the geometric ensemble average of microscopic units. 
An interpretation of $\alpha$ approaching $0$ could be that the limiting factor emerges purely from the growth rate without the need for a second growth-limiting parameter, in line with the previous crucial hypothesis of causality reversal stated above. One further simplification could be seen in equating the logarithm of the geometric mean with the logarithm of the median of the set of $x_i$ to obtain Eq. \ref{eq:medianGomp},
\begin{equation}
\frac{d}{dt}\ln{M(t)} = -\beta \ln{M(t)}.
\end{equation}

>>>>

\insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of deceased from the Coronavirus, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing an SIRD model (green) with a Gompertz model (yellow) for a variety of countries in the period Jan-May 2020. Both models are fit using non-linear least squares according to equations \ref{eq:GompertzODE} and \ref{eq:SIRD_D} in the text. Although the Gompertz curve can also be obtained with a simple linear fit using Eq. \ref{eq:GompertzODE2param}, it is here obtained using non-linear least squares to put both models on equal footing. The temporal evolution of the SIRD model is obtained from the Runge-Kutta algorithm, while the Gompertz has a closed form for its temporal evolution. Each plot is annotated with the basic reproduction number defined as $\mathcal{R}_0=\gamma\xi/\kappa$ \citep{carletti2020covid}. The $95^{th}$ percent confidence interval is shown in brackets and computed with the parametric bootstrap assuming a multivariate normal distribution on the parameters and 10000 generated bootstrap samples \citep{efron2012bayesian}. Fitting is done with Python-Scipy and a 6-day moving average of deaths. Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \cite{dong2020interactive}.}

@@

\section*{Comparing logistic and Gompertzian growth in the microscopic domain}
The remarkable observation that the log-transformed domain is the natural one merits closer study. First, juxtapose the logistic model with the Gompertz model,

@@

\section*{Unifying the SIR model with Gompertzian growth}
We now reconcile the SIR model with the Gompertz curve. 
Start with the networked SIR equation for the infected group,
\begin{equation}
\label{eq:SIRnetwork}
\frac{d x_i}{dt} = \beta (1-x_i)\sum_j{a_{ij}}x_j - \alpha x_i\quad \forall i,
\end{equation}

We then modify this equation in two ways: 1. Reverse the causality where the population of infected is now dependent on the population of the susceptible. This causality reversal is the crucial change based on the environmental stressor hypothesis rather than a communicable disease assumption. 2. Augment the transmission term to higher orders, which as we shall see is necessary to achieve Gompertzian growth.

Thus, Letting $s_i = 1 - x_i$ be the probability of being susceptible,

\begin{equation}
\label{eq:modSIR}
\frac{d x_i}{dt} = \beta x_i\sum_j{a_{ij}}(s_{j} + s^2_j/2 + ...) - \alpha x_i\quad \forall i,
\end{equation}

Furthermore, we will for the sake of simplicity assume all nodes in the network have exactly one neighbor and that there exists a unique path between all nodes,
\begin{equation}
\sum_i a_{ij} = 1 \quad \forall j.
\end{equation}

To build some intuition, we consider some simple scenarios. 
\begin{enumerate}
\item Let $x_j=1$, $\forall j \neq i$. In this scenario, Eq. \ref{eq:modSIR} becomes $\frac{d x_i}{dt} = - \alpha x_i$, implying as expected that the environmental stressor is gone, and the value of $x_i$ goes to zero. In the communicable disease paradigm, Eq. \ref{eq:SIRnetwork} becomes $\frac{d x_i}{dt} = \beta (1-x_i) - \alpha x_i$, and $x_i$ will growth as long as the right-hand side is positive.
\item Let $x_j \rightarrow 0$, $\forall j \neq i$. In this scenario, Eq. \ref{eq:modSIR} $\frac{1}{x_i}\frac{d x_i}{dt} \rightarrow \infty$, implying asymptotically infinite initial growth as the number of higher order terms increases. This behavior is in line with Gompertzian growth and is expected as growth is driven by the susceptible. The implication is that the environmental stressor has just started. In the communicable disease paradigm Eq. \ref{eq:SIRnetwork} becomes $\frac{d x_i}{dt} = - \alpha x_i$, and the value of $x_i$ goes to zero due to the absence of infectious people.
\end{enumerate}
It is important to note that both sets of equations only describe the paradigm they represent. Consequently, the set of equations for one paradigm will appear nonsensical in the other paradigm.

@@

\begin{equation}
\frac{d x_i}{dt} = -\beta x_i\sum_j{a_{ij}}\ln{(1 - s_j)} - \alpha = -\beta x_i\sum_j{a_{ij}}\ln{x_j} - \alpha x_i \quad \forall i,
\end{equation}
As we are interested in the aggregate macroscopic behavior, we take averages in the log-domain, exploit our setup where $\mathbf{A}$ is a single mapping from one node to another, and simplify to
\begin{equation}
\frac{d}{{dt}} \ln\left[\prod_i^N{x_i}\right]^{\frac{1}{N}} = -\beta\ln\left[\prod_i^N{x_i}\right]^{\frac{1}{N}} - \frac{\alpha}{N}.
\end{equation}
If $\alpha\rightarrow 0$, then this equation will exhibit Gompertzian growth in the geometric ensemble average of microscopic units. 
An interpretation of $\alpha$ approaching $0$ could be that the limiting factor emerges purely from the growth rate without the need for a second growth-limiting parameter, in line with the previous crucial hypothesis of causality reversal stated above. One further simplification could be seen in equating the logarithm of the geometric mean with the logarithm of the median of the set of $x_i$ to obtain Eq. \ref{eq:medianGomp},
\begin{equation}
\frac{d}{dt}\ln{M(t)} = -\beta \ln{M(t)}.
\end{equation}
We see that in a mathematically well-defined scenario, nodes in the network respond instantaneously and correlated to each other as under an environmental stressor.
